Reasonably Programmable Literal Notation

### 2017-10-26 pldi18/syntax-pldi18.tex
ported POPL submission to PLDI format and started doing some edits

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

% \newtheorem{theorem}{Theorem}[chapter]
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{condition}[theorem]{Condition}

%% Listings
\definecolor{mygray}{rgb}{0.75,0.75,0.75}
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray},
stringstyle=\ttfamily\color{red},
mathescape=false,escapechar=~,
numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=3pt, morekeywords=[1]{spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0,
aboveskip=4pt,belowskip=2pt,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{OliveGreen}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"},
}
\lstset{morecomment=[n]{/*}{*/}}

@@

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

@@

%% Title information
\title[Reasonably Programmable Literal Forms]{Reasonably Programmable Literal Forms}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

General-purpose programming languages typically provide literal notation for only a small number of standard data structures, like lists. This common design practice is decidedly \emph{ad hoc}: there are many other data structures for which literal notation might also be useful, e.g. finite maps, regexes, HTML data, SQL queries, syntax trees and chemical structures. Availability of literal notation is far from a superficial concern: studies suggest that literal notation can have a serious impact on software quality by subverting string encodings, which are vectors for injection attacks and cause other difficulties. %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
This paper develops, from type-theoretic first principles, a mechanism that decentralizes control over literal notation: \emph{typed literal macros (TLMs)}. TLMs give library providers programmatic control over the parsing and expansion of expressions and patterns of a flexible \emph{generalized literal form} at {a} specified type or parameteric family of types. % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding. 

@@

Compared to existing approaches, including syntax extension systems and macro-based quotation systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program. The system only needs to convey to clients (via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of spliced terms. Maintaining information about segmentation is necessary for the novel TLM hygiene mechanism, i.e. the mechanism that ensures that clients can reason abstractly about binding (previously, only unhygienic macro systems supported splicing of terms out of string literal bodies.) 

@@

This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, patterns and modules. This calculus is the  first detailed type-theoretic account of a hygienic term-rewriting macro system, of any design, for a language with these essential features of ML. We are integrating TLMs into Reason, an emerging alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.

@@

% \vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is a common practice to build in \emph{literal forms} that decrease the syntactic cost of constructing and pattern matching over values of a select few standard types. For example, in Standard ML (SML) and its cousins, a list expression literal like \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. We will return to the concept of context independence again.} Another example is found in Ur/Web's surface syntax, which extends Ur's surface syntax with expression and pattern literal forms for encodings of HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Fig. \ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2). 
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

This common language design practice is decidedly \emph{ad hoc} -- it is quite easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}. Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, set and map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web does not have); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although concerns about ``missing'' literal notation are easy to dismiss as superficial or frivolous, the reality is that literal forms, or the absence thereof, have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal forms for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of syntactic convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, file system paths, URLs and many other data structures. Literal forms, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web.}
\vspace{-8px}
\label{fig:urweb}
\end{figure}

Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with built-in literal forms. Instead, there has been considerable interest in mechanisms that allow library providers to define new literal notation on their own. For example, general syntax extension mechanisms, including Camlp4 \cite{ocaml-manual}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and others that we will return to in Sec. \ref{sec:existing-approaches}, can be used to define new literal forms. The problem that specifically motivates this paper is that these mechanisms can obscure program structure and are therefore \emph{unreasonable} for ``programming in the large'' \cite{DeRemer76}.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
Consider, for instance, the perspective of a programmer attempting to comprehend (i.e. reason about) the program text in Fig. \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason\footnote{\url{http://facebook.github.io/reason}} that has, hypothetically, been extended with a large number of new literal forms.

\begin{figure}[h]
\vspace{-10px}
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = [|(!R)@&{&/x!/:2_!x}'!R}|];
\end{lstlisting}
\vspace{-10px}
\caption{Syntax extension systems produce unreasonable program text.}
\vspace{-10px}
\label{fig:K-dialect}
\end{figure}

\noindent
The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its terse surface syntax \cite{Whitney:2001:LOR:376284.375783}. Of course, the programmer reading this program text may be unfamiliar with K's query syntax. Unfortunately for this programmer, there are no simple, abstract protocols for answering questions like:

@@

This paper formally details a more reasonable mechanism for defining new literal forms: \emph{typed literal macros (TLMs)}. TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) For example, Fig. \ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a generalized literal form delimited by \li{[|} and \li{|]} to express the ``unreasonable'' K query example from Fig. \ref{fig:K-dialect}. The applied TLM parses the literal body and generates an expansion statically, during a semantic phase called typed expansion, which generalizes the usual typing phase.  TLMs come equipped with uniquely powerful abstract reasoning principles. We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuitions -- the programmer can reason abstractly about Line 3 as follows:
\begin{enumerate}[noitemsep,nolistsep]
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, parses and determines the expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider is able to modularly establish that the syntax this TLM implements is deterministic.
\item \textbf{(Segmentation)} The output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. So by examining Fig. \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K query language (because they are in green.)
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes with a type annotation. %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = $kquery [|SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL|];
\end{lstlisting}
\vspace{-10px}
\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
\vspace{-12px}
\label{fig:K-tsm-example}
\end{figure}

The remainder of this document is organized as follows.
% \begin{itemize}
% \item 
Sec. \ref{sec:setsms} defines another {simple expression TLM} -- one that implements the Ur/Web-style HTML literal syntax from Fig. \ref{fig:urweb} -- and uses it as a running example to informally describe how TLMs work from start to finish.
% \item 
Sec. \ref{sec:sptsms} then briefly introduces {pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
% \item 
Sec. \ref{sec:ptsms} then adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that can take type and module parameters. Parameters serve two important purposes. First, they allow a TLM to operate not just at a single type, but over a type- and module-parameterized family of types. We give as an example a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type. Second, parameters lower the syntactic cost of the strict context independence discipline that TLMs enforce because they give expansions access to terms other than those explicitly spliced in to the literal body by the client, and because %This is particularly convenient for working with abstract types \emph{a la} ML. 
partial parameter application is possible. % Sec. \ref{sec:ptsms} culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
% \item
Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.
% \item 
Sec. \ref{sec:existing-approaches} compares TLMs to related work, including recent related work by \citet{TSLs,sac15}, and the long history of related work on macro systems.
% \item 
Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of the limitations of TLMs, along with some suggestions for future work.
% \end{itemize}

@@

\vspace{-4px}
\newcommand{\seTLMsSec}{Simple Expression TLMs (seTLMs)}
\section{\protect\seTLMsSec}
\label{sec:setsms}
\vspace{-2px}
We will start in this section by informally describing TLMs in Reason. It will suffice in this section to consider only \emph{simple expression TLMs (seTLMs)}, i.e. TLMs for abbreviating expressions of a single specified type. In particular, we will consider an seTLM named \li{$html} that implements Ur/Web style HTML literal notation for expressions of the \li{html} type that is defined in Fig. \ref{fig:html-type-def}.

@@

\vspace{-4px}
\subsection{TLM Definition and Application}
\vspace{-2px}
The definition of the \lstinline{$html} TLM is outlined in Fig. \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
 to distinguish them from variable identifiers. Every TLM definition also has a \emph{type annotation}, here \lstinline{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}. The TLM definition is in scope until the end of the enclosing scope. We will describe how TLM definitions are packaged into libraries in Sec \ref{sec:static-eval}.

>>>>

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
\documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

% \newtheorem{theorem}{Theorem}[chapter]
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{definition}[theorem]{Definition}
% \newtheorem{assumption}[theorem]{Assumption}
% \newtheorem{condition}[theorem]{Condition}

% \usepackage{titlesec}
% \titlespacing{\paragraph}{0pt}{10pt}{10pt}

\makeatletter
\renewcommand{\paragraph}{%
  \@startsection{paragraph}{4}%
  {\z@}{0.25ex \@plus 1ex \@minus .2ex}{-1em}%
  {\normalfont\normalsize\bfseries}%
}
\makeatother

%% Listings
\definecolor{mygray}{rgb}{0.75,0.75,0.75}
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray},
stringstyle=\ttfamily\color{red},
mathescape=false,escapechar=~,
numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0, xleftmargin=-10pt, 
aboveskip=4pt,belowskip=2pt,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{OliveGreen}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"},
}
\lstset{morecomment=[n]{/*}{*/}}

@@

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
% \citestyle{acmauthoryear}   %% For author/year citations

@@

%% Title information
\title[Reasonably Programmable Literal Forms]{Reasonably Programmable Literal Notation}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

General-purpose programming languages typically provide literal notation for only a small number of data structures, like lists. This design practice is unsatisfying because there are many other data structures for which literal notation might be useful, e.g. finite maps, regexes, HTML data, SQL queries, syntax trees and encodings of chemical structures. %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
This paper introduces \emph{typed literal macros (TLMs)}, which allow library providers to define new literal notation of nearly arbitrary form at a specified type or parameterized family of types. 
% TLMs give library providers programmatic control over the parsing and expansion of expressions and patterns of a flexible \emph{generalized literal form} at {a} specified type or parameteric family of types. % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding. 

@@

Compared to existing approaches, including direct grammar extension systems and prior  macro-based parsing systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic ambiguity, and TLM clients can reason abstractly, i.e. without examining the underlying expansion, about types and binding. The system only needs to convey to clients via secondary notation the inferred \emph{segmentation} of each literal, which gives the locations of spliced subterms. 
%Maintaining information about segmentation is necessary for the novel TLM hygiene mechanism, i.e. the mechanism that ensures that clients can reason abstractly about binding (previously, only unhygienic macro systems supported splicing of terms out of string literal bodies.) 

@@

This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, patterns and modules. This calculus is the first detailed type-theoretic account of a hygienic term-rewriting macro system, of any design, for a language with these essential features of ML, and the first literal macro system equipped with these essential reasoning principles. We are integrating TLMs into Reason, a popular alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.

@@

% \vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
\paragraph{Motivation} When designing the textual syntax of a general-purpose programming language, it is a common practice to build in \emph{literal forms} that decrease the syntactic cost of constructing and pattern matching over values of a select few standard types. For example, in Standard ML (SML) and its cousins, a list expression literal like \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. We will return to the concept of context independence again.} Another example is found in Ur/Web's textual syntax, which extends Ur's syntax with expression and pattern literal forms for encodings of HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2). 
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

This language design practice is decidedly \emph{ad hoc} -- it is quite easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}. Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, set and map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web does not have); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although concerns about the availability of literal notation are easy to dismiss as superficial or frivolous, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, file system paths, URLs and many other data structures. Literal forms, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure*}
\vspace{-2px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\vspace{-8px}
\label{fig:urweb}
\end{figure*}

Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with built-in literal forms. Instead, there has been considerable interest in mechanisms that allow library providers to define new literal notation on their own. For example, general syntax extension mechanisms, including Camlp4 \cite{ocaml-manual}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's type-specific languages (TSLs) \cite{TSLs}, and others that we will return to in Sec. \ref{sec:existing-approaches}, can be used to define new literal forms.

\paragraph{Problem} The problem that specifically motivates this paper is that these mechanisms obscure the formal structure of the program and thus make it difficult for programmers to reason about the program text directly. 
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''

Consider, for instance, the perspective of a programmer attempting to comprehend (i.e. reason about) the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason\footnote{\url{http://facebook.github.io/reason}} that has, hypothetically, been extended with some large number of new literal forms.

\begin{figure}[t!]
\vspace{-2px}
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = [|(!R)@&{&/x!/:2_!x}'!R}|];
\end{lstlisting}
\vspace{-10px}
\caption{Syntax extension systems produce unreasonable program text.}
\vspace{-10px}
\label{fig:K-dialect}
\end{figure}

The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its terse surface syntax \cite{Whitney:2001:LOR:376284.375783}. Of course, the programmer reading this program text may be unfamiliar with K's query syntax. Unfortunately for this programmer, there are no simple, abstract protocols for answering questions like:

@@

\paragraph{Approach}

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = $kquery [|SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL|];
\end{lstlisting}
\vspace{-10px}
\caption{TLMs make examples like the one from Figure \ref{fig:K-dialect} more reasonable.}
\vspace{-10px}
\label{fig:K-tsm-example}
\end{figure}

This paper formally details a more reasonable mechanism for defining new literal forms: \emph{typed literal macros (TLMs)}. TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) For example, Figure \ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a generalized literal form delimited by \li{[|} and \li{|]} to express the ``unreasonable'' K query example from Figure \ref{fig:K-dialect}. The applied TLM parses the literal body and generates an expansion statically, during a semantic phase called typed expansion, which generalizes the usual typing phase.  TLMs come equipped with uniquely powerful abstract reasoning principles. We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuitions -- the programmer can reason abstractly about Line 3 as follows:
\begin{enumerate}[noitemsep,nolistsep]
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, parses and determines the expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider is able to modularly establish that the syntax this TLM implements is deterministic.
\item \textbf{(Segmentation)} The output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K query language (because they are in green.)
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes with a type annotation. %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

The remainder of this document is organized as follows.
% \begin{itemize}
% \item 
Sec. \ref{sec:setsms} defines another {simple expression TLM} -- one that implements the Ur/Web-style HTML literal syntax from Figure \ref{fig:urweb} -- and uses it as a running example to informally describe how TLMs work from start to finish.
% \item 
Sec. \ref{sec:sptsms} then briefly introduces {pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
% \item 
Sec. \ref{sec:ptsms} then adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that can take type and module parameters. Parameters serve two important purposes. First, they allow a TLM to operate not just at a single type, but over a type- and module-parameterized family of types. We give as an example a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type. Second, parameters lower the syntactic cost of the strict context independence discipline that TLMs enforce because they give expansions access to terms other than those explicitly spliced in to the literal body by the client, and because %This is particularly convenient for working with abstract types \emph{a la} ML. 
partial parameter application is possible. % Sec. \ref{sec:ptsms} culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
% \item
Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.
% \item 
Sec. \ref{sec:existing-approaches} compares TLMs to related work, including recent related work by \citet{TSLs,sac15}, and the long history of related work on macro systems.
% \item 
Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of the limitations of TLMs, along with some suggestions for future work.
% \end{itemize}

@@

\vspace{-4px}
\newcommand{\seTLMsSec}{Simple Expression TLMs (seTLMs)}
\section{\protect\seTLMsSec}
\label{sec:setsms}
\vspace{-2px}
We will start in this section by informally describing TLMs in Reason. It will suffice in this section to consider only \emph{simple expression TLMs (seTLMs)}, i.e. TLMs for abbreviating expressions of a single specified type. In particular, we will consider an seTLM named \li{$html} that implements Ur/Web style HTML literal notation for expressions of the \li{html} type that is defined in Figure \ref{fig:html-type-def}.

@@

\vspace{-4px}
\subsection{TLM Definition and Application}
\vspace{-2px}
The definition of the \lstinline{$html} TLM is outlined in Figure \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
 to distinguish them from variable identifiers. Every TLM definition also has a \emph{type annotation}, here \lstinline{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}. The TLM definition is in scope until the end of the enclosing scope. We will describe how TLM definitions are packaged into libraries in Sec \ref{sec:static-eval}.
